Resolvable Ambiguity

### 2019-06-16 paper/paper.tex
EOD

\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{\mathbb{L}} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}

@@

This section briefly describes the theoretical foundations we build upon. Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation. Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}. Finally, Sections~\ref{sec:preliminaries-trees} and \ref{sec:preliminaries-ambiguity} describe trees and ambiguity, the former with some non-standard notation and the latter with a slightly wider definition than normal.

@@

A context-free grammar $G$ is a tuple $(\NT, \T, P, S)$ where:

\begin{itemize}
\item $\NT$ is a set of non-terminals.
\item $\T$, disjoint from $\NT$, is a set of terminals.
\item $P$, a subset of $\NT \times (\NT \cup \T)^{*}$, is a set of productions.
\item $S \in \NT$ is the starting non-terminal.
\end{itemize}

\noindent A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$. The set of words recognized by $G$ is written $L(G)$, and is said to be the word language of $G$.

@@

A nondeterministic finite automaton (NFA) is a tuple $(Q, \Sigma, \delta, q_0, F)$:

\begin{itemize}
\item A finite set of states $Q$.
\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
\item A transition function $\delta: Q \times \Sigma -> 2^Q$.
\item An initial state $q_0 \in Q$.
\item A set of final states $F \subseteq Q$.
\end{itemize}

\noindent A successful run is a sequence of states $r_0, r_1, \ldots, r_n$ and a word $a_0a_1\ldots a_n$ such that:

@@

A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to.

A pushdown automaton extends a finite automaton with a stack, and lets each transition push or pop symbols from it. Formally, a nondeterministic pushdown automaton is a tuple $(Q, \Sigma, \Gamma, \delta, q_0, F)$:

\begin{itemize}
\item A finite set of states $Q$.
\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet, including a special bottom of stack symbol $\bot$.
\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma^{*}}$.
\item An initial state $q_0 \in Q$.
\item A set of final states $F \subseteq Q$.
\end{itemize}

\noindent A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma*$.

@@

\begin{itemize}
\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all.
\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack.
\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack (i.e., this transition cannot be taken if $g$ is not on top of the stack).
\end{itemize}

@@

A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton. A visibly pushdown automaton is a pushdown automaton where the input alphabet $\Sigma$ can be partitioned into three disjoint sets $\Sigma_c$, $\Sigma_i$, and $\Sigma_r$, such that all transitions in the automaton has one of the following three forms:

\begin{itemize}
\item $p \xrightarrow{c, +s} q$, where $c \in \Sigma_c$.
\item $p \xrightarrow{i} q$, where $i \in \Sigma_i$.
\item $p \xrightarrow{r, -s} q$, where $r \in \Sigma_r$.
\end{itemize}

@@

This gives us some nice properties. Of particular relevance to this paper are the following two points:

\begin{itemize}
\item If two visibly pushdown automata have the same partitions $\Sigma_c$, $\Sigma_i$, and $\Sigma_r$, then we can construct a product automaton that simulates two simultaneous runs through both automata. This product automaton has the same input alphabet partitions, each state is a pair of states (one from each automaton), and each stack symbol is a pair of stack symbols (one from each automaton).
\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run, none are redundant.
\end{itemize}

@@

Trees generalize words by allowing each terminal to have more than one successor. The successors have an associated ordering, i.e., in the presence of multiple successors, one is considered to be first, one second, etc. Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors. This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed. In this case, the sequence of successors to a single terminal tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.

Our later results and properties are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature. We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves, to give a more natural definition of $\yield$.

@@

\begin{itemize}
\item $\NT$ is a set of (zero-arity) non-terminals.
\item $\T$ is a set of zero-arity terminals, used as leaves.
\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
\item $P$ is a set of productions, where a production has the form $N -> x(r)$ for $N \in \NT$, $x \in X$, and $r$ a regular expression over $\NT$ and $\T$.
\end{itemize}

\noindent A tree $t$ is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$. The set of trees recognized by $T$ is written $L(T)$\footnote{All word grammars will be named $G$ (possibly with a subscript), while all tree grammars will be named $T$ (possibly with a subscript), lessening the risk of confusion between the two cases.}, and is said to be the tree language of $T$.

$\yield$ then, is the sequence $w \in \T^{*}$ of terminals $t \in \T$ obtained by a left-to-right traversal of a tree. Informally, it is the flattening of a tree after all internal nodes have been removed.

The standard definition of ambiguity, given a grammar $G$ and its parse trees $T_G$, is as follows:

\begin{definition}
A word $w \in L(G)$ is ambiguous if:
  \[ \exists t_1, t_2 \in L(T_G).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2) \]
\end{definition}

\noindent For this paper, we widen the definition slightly to the following:

\begin{definition}
  A word $w \in L(G)$ is ambiguous relative a tree grammar $T$ and a function $\parse : L(G) -> 2^{L(T)}$ if:
  \[ |\parse(w)| > 1 \]
\end{definition}

\noindent where $2^S$ is the powerset of $S$ and $|S|$ is the cardinality of $S$. The standard ambiguity definition is regained by choosing

\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(T_G) \}\ \]

\section{Parse-time Disambiguation}

We begin this section with some motivation, then list our definition of \emph{resolvable ambiguity} and what disambiguation we will consider, and finally an alternative definition of a word, which will be useful in later sections.

We can divide the productions present in a programming language grammar in two groups: those that are semantically important, and those that are semantically \emph{un}important. The former group covers most productions, while the latter contains, e.g., parentheses used for explicit grouping. If programs were written directly as syntax trees then the latter would be unnecessary; two syntax trees that differ only by parentheses are semantically the same.

Thus we wish the output of parsing to be a syntax tree consisting entirely of semantically important productions. This distinction is useful to make, because it allows us to decouple \emph{what} we want to be expressible and \emph{how} it is to be expressed.

@@

With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}. Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label. A \emph{mark} is used for disambiguation, and forbids certain productions from taking the place of the marked non-terminal. To lessen clutter, we will write $E_\emptyset$ as $E$. As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example. In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a child of a multiplication, enforcing conventional precedence.

From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:

\begin{itemize}
\item $G_D$ represents a word language describing all semantically distinct programs.
\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation.
\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
\end{itemize}

\begin{figure}

  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T_D$, the parse trees of $G_D$.}
    \label{fig:running-example-generated:t}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \midrule
      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T'_D$, the parse trees of $G'_D$.}
    \label{fig:running-example-generated:t-prime}
  \end{subfigure}

@@

\begin{figure*}[ht]
  \begin{tikzpicture}[every node/.style={rectangle,draw}]
    \node (a) [draw=white] at (0,5) {$((1) + (2)) + (3)$};

>>>>

\newcommand{\support}{\mathit{Supp}}
\newcommand{\NT}{V} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{L} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}

@@

This section briefly describes the theoretical foundations we build upon. Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation to make later sections easier to read. Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}. Finally, Sections~\ref{sec:preliminaries-trees} and \ref{sec:preliminaries-ambiguity} describe trees and ambiguity, the former with some non-standard notation and the latter with a slightly wider definition than normal.

\subsection{Bags} \label{sec:bags}

A \emph{bag} (alternatively known as a \emph{multiset}) is a generalization of a set, it is an unordered collection where each element may appear multiple times. The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$. A bag $A$ is included in another bag $B$, written $A \subseteq B$, iff $\forall x.\ m_A(x) \leq m_B(x)$. The underlying set of elements of a bag $B$ is called the \emph{support} of $B$, and is given by $\support(B) = \{ x \mid m_B(x) > 0 \}$.

\subsection{Regular Expressions} \label{sec:preliminaries-regexes}

A regular expression $r$ over alphabet $\T$ is defined inductively:

$$r ::= \epsilon \mid a \mid r \cdot r \mid r + r \mid r^{*}$$

\noindent where $a \in \T$. The language of a regular expression $r$ is given by $L(r)$:

$$
\begin{array}{r@{\ =\ }l}
  L(t) & \{a\} \\
  L(\epsilon) & \{\epsilon\} \\
  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \\
  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
\end{array}
$$

\noindent We will denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.

@@

A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is Kleene-star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.

 A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$. Such a sequence is called a \emph{derivation}. The set of words recognized by $G$ is written $L(G)$\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.

The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{left-most derivations}. A left-most derivation is a derivation where the non-terminal being replaced is always the left-most one.

\begin{definition}
A word $w \in L(G)$ is ambiguous if there are two distinct left-most derivations of that word.
\end{definition}

@@

A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$ ; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.

A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that:

@@

A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to. NFAs and DFAs have the same expressive power as regular expressions, i.e., for every regular expression there is an NFA and a DFA both reconizing the same language, and vice-versa. % TODO: probably reference for this

A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols. Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states. $\lambda$ essentially means ''ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).

A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma^{*}$.

@@

\begin{itemize}
\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all, i.e., $\delta(p, a, \lambda) \supseteq \{(q, \lambda)\}$.
\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack, i.e., $\delta(p, a, \lambda) \supseteq \{(q, g)\}$.
\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack, i.e., $\delta(p, a, g) \supseteq \{(q, \lambda)\}$.
\end{itemize}

@@

A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton. A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton has one of the following three forms:

\begin{itemize}
\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$.
\item $p \xrightarrow{i} q$, where $i \in \T_i$.
\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$.
\end{itemize}

@@

The partition names stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}. We will instead primarily use $\T_c$ and $\T_r$ for balanced parentheses.

This partitioning gives us some useful properties. Of particular relevance to this paper are the following two points:

\begin{itemize}
\item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}. Intersection in particular is given by a product automaton, i.e., given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
  $$
  \begin{array}{r@{\ =\ }ll}
    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \text{where } c \in \T_c \\
    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \text{where } i \in \T_i \\
    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \text{where } r \in \T_r \\
  \end{array}
  $$

\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run, none are redundant. Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
\end{itemize}

@@

Trees generalize words by allowing each terminal to have multiple ordered successors, instead of just zero or one. Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors. This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed. The sequence of successors to a single terminal in an unranked tree tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.

The results and properties presented in this paper are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature. We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves. Since we will use unranked trees to represent parse trees, the former will represent terminals from the parsed word, while the latter represent terminals introduced as internal nodes.

@@

\begin{itemize}
\item $\NT$ is a set of (zero-arity) non-terminals.
\item $\T$ is a set of zero-arity terminals, used as leaves.
\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
\item $P$ is a set of productions, a finite subset of $\NT \times X \times \regex(\T \cup X)$. We will write a production $(N, x, r)$ as $N -> x(r)$.
\end{itemize}

\noindent A tree $t$ (containing only terminals from $\T$ and $X$) is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$. The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.

Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree. Informally, it is the flattening of a tree after all internal nodes have been removed.

\section{Resolvable Ambiguity} \label{sec:resolvable-definition}

This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to some more standard concepts in formal languages.

Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense. We will call this meaning an \emph{interpretation} of a word. Note that a single word may have multiple interpretations. As such, we will define a language as:

\begin{itemize}
\item An alphabet $\T$.
\item A set of interpretations $T$.
\item A function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
\end{itemize}

\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees. In such a language, $\parse(1 + 2 * 3)$ would produce a set containing the two ASTs in Figure~\ref{fig:arith-example}:

\begin{figure}[t]
\begin{tabular}{cc}
  \Tree [.{$+$}
    1
    [.{$*$}
      2
      3 ] ] &
  \Tree [.{$*$}
    [.{$+$}
      1
      2 ]
    3 ] \\
\end{tabular}
\caption{The two allowable interpretations of $1 + 2 * 3$ in an arithmetic language without precedence.}
\label{fig:arith-example}
\end{figure}

\noindent We further classify each word $w \in \T^{*}$ depending on the result of $\parse$:

\begin{definition}
  A word $w \in \T^{*}$ is:
  \begin{itemize}
  \item not in the language if $\parse(w) = \emptyset$.
  \item unambiguous if $\exists t \in T.\ \parse(w) = \{t\}$.
  \item ambiguous if $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$.
  \end{itemize}
\end{definition}

\noindent We can connect these cases to the classical definition of a language:

\begin{itemize}
\item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ that relates words to their left-most derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

\noindent A \emph{resolvably} ambiguous word is a word where all its interpretations can be written in an unambiguous way, or more formally:

\begin{definition}
  A word $w \in \T^{*}$ in a language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-word}
\end{definition}

\noindent The example in Figure~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 * 3)$ and $(1 + 2) * 3$ are unambiguously interpreted as the first and second tree, respectively.

Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, or more formally:

\begin{definition}
  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall w \in \T^{*}.\ \forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-language}
\end{definition}

\noindent We can now state a few things:

\begin{itemize}
\item An unambigous word $w$ is trivially resolvably ambiguous, since its only interpretation $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$). The set of resolvably ambiguous words is thus a superset of the unambiguous words.
\item If a given interpretation $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous. In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
\end{itemize}

\noindent The second point suggests that resolvable ambiguity is only an interesting property if an element of $T$ does not uniquely identify an element of $\T^{*}$. Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual interpretation. Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace. In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the interpretation.

We thus propose to loosen the common ''no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous. However, merely having an arbitrary function $\parse$ gives us very little to work with, and no way to decide whether the language it defines is resolvably ambiguous or not. The remainder of this paper will thus consider $\parse$ functions defined with a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, that gives us some decidable properties.

Before introducing this formalism however, we introduce the two central problems we consider in this paper:

\begin{description}
\item[Static resolvability.] Given a language defined by $\parse$, determine if it is resolvably ambiguous.
\item[Dynamic resolvability.] Given a language defined by $\parse$ and a word $w$ such that $\parse(w) \neq \emptyset$, determine if $w$ is resolvably ambiguous.
\end{description}

\noindent Our main concern is producing decision procedures for these problems. As such, we define soundness and completeness for the static problem:

\begin{definition}
  A decision procedure $f$ solving the static resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse) = \mathit{resolvable}$ implies that the language given by $\parse$ is resolvably ambiguous, and $f(\parse) = \mathit{unresolvable}$ implies that the language given by $\parse$ is not resolvably ambiguous.
  \item Complete if $f$ terminates and produces a result for all possible inputs $\parse$.
  \end{itemize}
  \label{def:static-procedure}
\end{definition}

\noindent Similarly, for the dynamic problem:

\begin{definition}
  A decision procedure $f$ solving the dynamic resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
  \item Complete if $f$ terminates and produces a result for all possible inputs $\parse$ and $w$.
  \end{itemize}
  \label{def:dynamic-procedure}
\end{definition}

\section{Parse-time Disambiguation} \label{sec:parse-time-disambiguation}

This section describes our chosen language definition formalism, and motivates its design.

The primary purpose of this formalism is, as described in the previous section, to produce a $\parse$ function, i.e., to describe a word language and assign one or more interpretations to each word. The interpretations will be unranked trees, intended to be somewhat reminiscent of the abstract syntax trees used in most compilers. Section~\ref{sec:resolvable-definition} suggests that $\parse$ discard some information, to enable the resolution of some ambiguities; we here choose to discard grouping parentheses.

With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}. Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label. Also note that the right-hand side of a production is here a regular expression, rather than the theoretically simpler sequence used in a context-free grammar. Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal. To lessen clutter, we will write $E_\emptyset$ as $E$. As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example. In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication. By ''direct child'' we mean ''without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.

@@

From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$. Technically, only $G'_D$ and $T_D$ are required, $G'_D$ is used as the defined word language and $T_D$ as the interpretations, but the remaining two grammars help the presentation.

\begin{itemize}
\item $G_D$ represents a word language describing all semantically distinct programs.
\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation (i.e., the result of marks).
\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
\end{itemize}

\noindent Figure~\ref{fig:running-example-generated} contains the four grammars generated from our running example in Figure~\ref{fig:running-example-definition}. The context-free grammars are produced by a rather standard translation from regular expressions to CFGs, while the primed grammars get a new non-terminal per distinctly marked non-terminal in $D$, where each new non-terminal only has the productions whose label is not in the mark. For example, the non-terminal $E_{\{a\}}$ in Figure~\ref{fig:running-example-generated:t-prime} has no production corresponding to the $a$ production in Figure~\ref{fig:running-example-definition}.

\begin{figure}
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T_D$, the parse trees of $G_D$.}
    \label{fig:running-example-generated:t}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \midrule
      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T'_D$, the parse trees of $G'_D$.}
    \label{fig:running-example-generated:t-prime}
  \end{subfigure}

@@
